### 2021-07-02 ack.tex
got another intro figure constructed.

Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way. Tyler cannon. . .

Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,

Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.

Andy Osadetz, Josh Plankeel,

My partner Mary for putting up with the late nights.

>>>>

Attending university long enough to earn a PhD is an unbelieveable priviledge, and I only got here with an incredible amount of guidance and support.

Most of all, to mom, dad, Kelsey, Kim, and Gugs, you've formed all of the best parts of my personality, and it's impossible to explain how thankful I am for that.

Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on. 
I cannot imagine a better example of work ethic, open-mindedness, and devotion. I hope I picked a few things up.

Dad, I've watched you solve an insane range of problems without hesitation, from worn out brake pads to broken arms.
By age 10 I must have helped with copper plumbing, woodwork, car repair, home wiring, and tree felling.
I had personal tours of West Virginia's backroads and civil engineering works, and I learned more about science paddling the South Branch with you than I ever did in school.
To you, water currents, fish habits, electrons, hellgrammites, engines, heat, and radio waves are all deeply important, and any man-made problems are easy enough to solve with time and the right tools.
Thanks for sharing the perspective of a problem solver and knowledge seeker! And of course so much more.

Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking you for help every couple of paragraphs. You made me good at reading and taught me to enjoy it, which is the foundation of all my learning. Plus, we had a lot of hilarious times. I know it goes without saying, but thanks.

Gugs, your approach to life is clean and selfless, and I cannot imagine a single way you could have been more reliable, stable, or invested in my development.
I'll never be able to thank you enough!
One day I'll take up your 4am wakeups and adopt your completely measured and calm responses to life's challenges. Until then, I'll keep thinking about I'd ideally do in terms of what you always do, and if we're lucky I might even occasionally pull it off.

Another family member who deserves special mention is my great aunt Alice. Thanks so much for paying for my undergraduate education. We hardly met, but we're family, and that was enough. It really means a lot.

Thanks to my early educators who put in the extra care to keep me on track, Jonathan Ramey, Linda Afzhalirhad, Brandon Willard, Linda Mendez, Marty Ojeda, Phyllis Adkins, and Red Hamlin among many others.
Thanks also to the later ones at Southern WV Community College (SWVCC) and West Virginia University (WVU) whose lessons fostered my way of thinking and provided me with many of the tools I used for the research in this thesis. At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble. Mindy in particular transformed my view of the world into one based on mechanics, and provided me with a deep understanding of general physics and calculus. Everyone there always encouraged me to continue my education. You all made SWVCC a personal transformation and an unbelievable privilege.

At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic. Alan and Tudor, thanks for providing my first opportunities in research. Leo, you are the key person in my intellectual development. You inspired me to learn mechanics, statistical physics, and mathematical physics at a high level, and I can't thank you enough. Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr. Golubovic would like this one." I wish I had taken more advantage of the opportunity to speak to you outside of classes. I'll feel deeply successful if I can one day become as inspiring an educator as you are.

To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Andrew Rice, Andrew Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times. I'll never forget Scott's quaternion obsession, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind). I have yet to experience teamwork and shared exploration like we had in the undergrad lounge. I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, and that they have some Andrew Rice imitator to flip some cars for em (mike jones). Because of you all, WVU was magic for me.

Now we come to Marwan Hassan. About five years ago now I wandered into Marwan's office saying I liked rivers and wanted to describe sediment movement with statistical physics. I realize now when I said ``statistical physics" your eyes lit up with thoughts of Einstein, Nakagawa, Yano, Furbish, Hubbell, and all of the names I now know well. I was lucky to use the right trigger words.
Marwan, thanks for taking the risk and taking me on from outside the field, for all of the guidance, for loaning me so many papers and books, giving me freedom to explore ideas, and most of all for showing me that it's ok to be human in science. At this point I also want to thank Yusuf for trying your patience to its limits over a long enough period so that things didn't seem so bad once I came along for this short time together.

My friends in geography have also been great. Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, David, Jiamei, Xingyu, Niannian, and Emma, plus the outsiders, Will, Dave, Rose, and Anya. In particular, thanks to Shawn for all of the chats and shared speculations, and Matteo and Conor -- friendships forged through suffering! We'll have to keep in touch.

Thank you Rui Ferreira and Brett Eaton!
Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics. Your kindness and wide-ranging interests beyond hydraulics are inspiring. Thanks for the thoughtful comments on my PhD. I'm not forgetting what I owe you! 
Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand modelling in light of the real world and its complexity. Although I'm not sure if I understood your lessons as well as you would have liked, I have a list of your favorite books on my computer, and they're a first task for as soon as I defend. Maybe they'll bring me closer to your way of thinking.

I'd be amiss not to mention David Furbish. Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation. At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes". Now I have a thesis on it. Later on, David recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social repercussions of disinformation, and a whole range of questions including the meaning of probability in landscape evolution, probability as being just as fundamental as mass, momentum, and energy, and fringe ideas in cosmology. These are my people! David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me hop on the spacecraft.

Among UBC Geographers, thanks especially to Nina Hewitt! It was always a pleasure and maybe a sigh of relief to TA your courses, which are so uniquely organized and student-focused. UBC Geography is a much greater department with you in it. I'm so thankful to have been associated with your influence on the undergrads. I'll be happy if I can teach like you one day!

UBC Geography is successful because it has excellent staff, Vincent Kujala, Sandy Lapsky, Bret Peterson, Suzanne Lawrence, Connie Cheung, Jose Aparicio, thank you all so much! 
Vincent showed me the magic of bash, Sandy handled conference expenses and got me back in an office during the pandemic when my productivity was floundering, Bret put out the tech fires, Suzanne organized the paychecks and appointments, and Jose handled every online teaching challenge. You all saved me a ton of time, and I really appreciate it.

The kind, open-minded, and collaborative atmosphere I experienced here is clearly a legacy from the former generation of UBC's physical geographers, some whom I had the pleasure of meeting and learning from, like Mike Church and Olav Slaymaker, and others whom I didn't. Thanks Mike and Olav for your inspiring seminars and the academic environment you cultivated.

Finally, Mary! My PhD has not been great in terms of work-life balance. Thanks for sticking it out, putting up with my late bedtimes, my early wakeups, and all of those social events I didn't come to. It's time! I'm done. The pandemic also seems to be winding down. We can finally get to Aus! I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella. After that, we're a step closer to settling somewhere and determining once and for all whether our future cat will be white or orange.

Completing a PhD looks like a personal achievement, but it's really not.
It's an expression of 30 years of influence and guidance from other people at every stage.
I'm kinda tired because it's the end. Maybe because of it, I find it easy to drift into daydreams on the parts of my life I've really loved, Watoga and Williams River with my family, reading encyclopedias in Gugs' living room, lining up just left of center with my dad and bracing on the paddle for the hit, nighttime fireflies under my great grandpa's tree, and sliding down rocks at Kumbrabow with my sisters. The daydream shifts to years later, as I flip on the light in the physics lounge and surprise Megat who was asleep on the couch, I finish a problem set with friends, discuss Tsujimoto with Marwan (``those guys did everything!"), or just walk around Kitsilano with Mary at the edge of dark. It's all one thing! Thank you all for every part of it.

### 2021-07-02 ch1intro.tex
got another intro figure constructed.

Landscapes evolve when water, wind, and ice, driven by gravity, erode the uplifted relics of forces from below Earth's surface.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands as they grind them to sand through a machine of denundation.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.

In this context river geomorphology as a scientific discipline is shifting toward more quantitative methods that enable concrete predictions about the natural world.
Sediment transport in river channels is especially amenable to this quantitative approach since it is basically the result of fluid and granular physics \citep{Frey2011}.
Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains.
When particles are coarse as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface. In bedload transport conditions, fluid turbulence and the irregular bed surface inherit special control over sediment dynamics.

Bedload transport exerts particular influence over channel morphology and stability \citep{Church2006,Hassan2008,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Zimmerman2007,Comiti2012, McKenzie2018, Eaton2020}.
A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
Unfortunately, existing approaches to compute the bedload flux are inadequate, with predictions regularly deviating from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012, Dhont2018}.
Considering that this is despite well over a century of research into bedload transport \citep{Gomez1991}, we can conclude that new research approaches are sorely needed \citep{Ancey2020a,Ancey2020}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Strom2008}, which encourages further transport in a positive feedback \citep{Ancey2008,Heyman2016,Lee2018}.
Bedload fluxes are thus in a cyclical feedback with their controls in the details of turbulence and bed organization \citep{Jerolmack2005}.

The approach taken in this thesis is to consider the bedload flux as an aggregate result of many individual transported grains.
This departs from the traditional strategy of correlating transport rates to the mean flow and sedimentary characteristics, which considers the granular phase as a continuum \citep{Gilbert1917,MeyerPeter1948}.
Since the trajectories of individual grains are governed by Newtonian mechanics, this alternative perspective provides traction on the problem.
This Newtonian approach is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time, in part due to the turbulent flow and the erratic interactions of moving particles with the bed.
To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
These efforts build on an an earlier literature to provide more realistic descriptions of the trajectories of individual grains and the bedload fluxes they generate.

The thesis begins with a review of the most relevant pieces of this earlier literature on which my research builds. This review rephrases works when necessary to indicate common themes and to show continuity with my own research which follows in the subsequent chapters.

@@

A basic (but not simple) problem in sediment transport modelling is to predict the downstream movement of an individual particle.
At first glance, this seems an elementary problem in general physics, but challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not known \citep{Furbish1997}.
Downstream movement is resisted by frictional collisions between moving particles and the bed.
Since the bed is a granular surface, its geometry is very difficult to characterize \citep{Gordon1972}, and the outcome of collisions varies from one to the next \citep{Sekine1992}
Entrainment and deposition provide an additional layer of complexity.
Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and enable their deposition \citep{Charru2004}.
These pockets are not permanent shelter, because rearrangement of the surrounding bed by conventional entrainment or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if the pocket geometry is not disturbed \citep{Valyrakis2012,Celik2014}. 
As a result, particles generally alternate through sequences of entrainment and deposition.
Ultimately, particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}. These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}.

Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales.
\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize this complexity.
Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
The local range refers to the period of motion between subsequent interactions with the bed, when the particle accelerates downstream within the flow, the intermediate range reflects particle motions through sequences of collisions, and the global range refers to particle motion between subsequent entrainment and disentrainment events.
\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried. 
This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize transport models of individual grains.

\subsection{Motivation: tracers and basic understanding}
The original motivation to understand individual particle motions was probably to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common measurement technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion in the stream to calculate the overall sediment flux \citep{Yano1969,Nakagawa1976,Hassan2017}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}. This means it is not clear how measured tracer velocities can be mapped back to bulk sediment fluxes, exposing a need for further research into how exactly particle motion characteristics evolve through time.

@@

Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest, but it is clearly oversimplified from the reality of particle transport, and a number of extensions and improvements have been made in the 85 years since its development.

>>>>

Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes uplifted by tectonics.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.

In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005}.
Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}. In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.

Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
Given that this problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}. The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018}.
Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values.

The approach taken in this thesis is to consider bedload transport as an aggregate result of many individual transported grains whose movements have probabilistic characteristics.
This departs from the traditional strategy of correlating transport rates to the average flow and sedimentary characteristics \citep{MeyerPeter1948, Parker1990, Wilcock2001}.
Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings reliable tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and the body of earlier literature which inspired my approach.
To provide context and motivation, the thesis begins by reviewing this literature, rephrasing earlier work as necessary to indicate common themes and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of an individual particle.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Furbish1997}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the outcomes of these encounters are variable \citep{Sekine1992}.

Entrainment and deposition provide an additional layer of complexity.
Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause their deposition \citep{Charru2004}.
These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2012,Celik2014}. 
As a result, particles generally alternate through sequences of entrainment and deposition, but determining precisely when these events occur is impractical.

Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}, 
These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Church1994,Ferguson2002}.

Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years, and each of these processes eludes precise description.

\citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
Nikora et al divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decellerate. 
The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017,Pretzslav2021}.
This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}. This obscures the mapping between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.

@@

Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest. Despite its success, it oversimplifies particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
